Installing packages:

```{r}
# install.packages("chron") # for working with dates / times
# install.packages("naniar") # tools for exploring & handing missing data
# install.packages("tidytuesdayR") # used to download TidyTuesday data
# install.packages("tidycensus") # an R package that allows users to interface with a select number of the US Census Bureau’s data APIs and return tidyverse-ready data frames
# install.packages("ggridges") # {ggplot2} extension for creating ridgeline plots
# install.packages("gghighlight") # {ggplot2} extension for highlighting geoms
# install.packages("ggbeeswarm") # {ggplot2} extension for creating categorical scatter (violin point) plots
# install.packages("see") # {ggplot2} extension for model visualization (we'll be using it for it's geom, geom_violindot())
# install.packages("scales") # provides the internal scaling infrastructure used by ggplot2, and gives you tools to override the default breaks, labels, transformations and palettes (installed automatically with {ggplot2} or {tidyverse})
```

# Source Key by running:

`source(here::here("week3", "KEYS.R"))`

# Choosing the right graphic form

Exercise: want to compare unemployment rates of 5 countries A, B, C, D, E

Hierarchy of elementary of perceptual tasks - viewer performs a mental-visual tasks to extract the values of real variables represented on most graphs - successful charts are constructed higher in the hierarchy\
- Bad at extracting exact values: hue / shade ; bad at looking at area - Bette at perceiving rectangles than circles

Caveats: doesn't include maps - lower scale methods can be appropriate when revealing **general patterns**

PURPOSE!!

How best to represent your data to serve purpose?

Things to think about...
1. Tasks you want to enable or messages you want to convey
2. Number of variables / data points / data types

Messages 
- primary / secondary insights of graphic 
- identify elemtnary perceptual tasks

# Visualizing Distributions 

Visualizing the SPREAD
- histograms
- density
- ridgeline plots
- box plots
- violin plots 

# Set up 
```{r}
# load packages 
library(tidyverse)
library(chron) # date time stuff
library(naniar) # missing data

# import data ----
mko <- read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-sbc.2007.17&entityid=02629ecc08a536972dec021f662428aa")
```

## Wrangle
```{r}
mko_clean <- mko %>% 
  select(year, month, day, decimal_time, Temp_bot) %>%   # select columns
  unite(date, # name of new column
        year, month, day, # columns to combine
        sep = "-", # separator
        remove = FALSE) %>%  # keep the columns we united
  mutate(time = times(decimal_time)) %>% 
  unite(date_time,
        date, time, 
        sep = " ") %>% 
  mutate(date_time = as.POSIXct(date_time, "%Y-%m-%d %H:%M:%S", # set to date time object
                                tz = "GMT"), # timezone
         year = as.factor(year),
         month = as.factor(month),
         day = as.numeric(day)) %>% 
  mutate(month_name = as.factor(month.name[month])) %>%  # index using the number in the month column 
  naniar::replace_with_na(replace = list(Temp_bot = 9999)) # assign values we want to replace with NA values

# look at structure using str()
```

# Plotting Histograms 
- shows the frequency of values in each range
```{r}
mko_clean %>% 
  mutate(month_name = factor(month_name, # assign column of interest
                             levels = month.name)) %>%  # arrange the months the same as this
  ggplot(aes(x = Temp_bot, fill = month_name)) +
  geom_histogram(position = "identity", # each bar on the x-axis
                 alpha = 0.5 # make transparent 
                 ) 

mko_clean %>% 
  mutate(month_name = factor(month_name, # assign column of interest
                             levels = month.name)) %>%  # arrange the months the same as this
  ggplot(aes(x = Temp_bot)) +
  geom_histogram(position = "identity" # each bar on the x-axis
                 ) +
  facet_wrap(~month_name)

# feweer groups =======
mko_clean %>% 
  mutate(month_name = factor(month_name, # assign column of interest
                             levels = month.name)) %>%  # arrange the months the same as this
  filter(month_name %in% c("April", "June", "October")) %>% 
  ggplot(aes(x = Temp_bot, fill = month_name)) +
  geom_histogram(position = "identity", # each bar on the x-axis
                 color = "black",
                 alpha = 0.5)
```
# Density plots
- proportion of values in each range 
- area under the curve = 1
```{r}
mko_clean %>% 
  mutate(month_name = factor(month_name, # assign column of interest
                             levels = month.name)) %>%  # arrange the months the same as this
  filter(month_name %in% c("April", "June", "October")) %>% 
  ggplot(aes(x = Temp_bot, fill = month_name)) +
  geom_density(position = "identity", # each bar on the x-axis
                 color = "black",
                 alpha = 0.5,
               adjust = 0.5) + # adjusts the smoothing (1 is the most smooth)
  scale_fill_manual(values = c("midnightblue", "orange", "pink"))
```
## hist vs density 
```{r}
dummy_data <- data.frame(value = c(rnorm(n = 100, mean = 5),
                                    rnorm(n = 200, mean = 10)),
                         group = rep(c("A", "B"),
                                     times = c(100, 200)))

# hist =====
ggplot(data = dummy_data, aes(x = value,
                       fill = group)) +
  geom_histogram(position = "identity", alpha = 0.7) +
  geom_rug(aes(color = group), alpha = 0.7)

# density =====
ggplot(data = dummy_data, aes(x = value,
                       fill = group)) +
  geom_density(position = "identity", alpha = 0.7) +
  geom_rug(aes(color = group), alpha = 0.7)
```
## ridgeline

```{r}
ggplot(mko_clean, 
       aes(x = Temp_bot, y = month_name, 
           fill = after_stat(x))) + # fill using the variable you assigned as x
  ggridges::geom_density_ridges_gradient(rel_min_height = 0.001, # length of tails
                                         scale = 3) + # overlap
  scale_y_discrete(limits = rev(month.name)) + # reorder months from January to December
  scale_fill_gradientn(colors = c("midnightblue", "lightblue", "white", "orange", "darkred"))

# can add jitter points to see sample size
```
## Boxplots
```{r}
ggplot(mko_clean, 
       aes(x = month_name,
           y = Temp_bot,
           fill = month_name)) +
  geom_boxplot() +
  scale_x_discrete(limits = rev(month.name)) + # reorder by month name
  coord_flip() +
  gghighlight::gghighlight(month_name == "October") + # highlight just one month
  theme(
    legend.position = "none" # remove legend
  )

# jitter and dodge
library(palmerpenguins)

penguins %>% 
  mutate(year = as.factor(year)) %>% 
  ggplot(aes(x = species, y = body_mass_g, color = year)) +
  geom_boxplot() +
  geom_point(alpha = 0.5, 
             position = position_jitterdodge(jitter.width = 0.2)) +
  coord_flip()
```

## violin plots 
- distributino of a numeric variable for one or several groups 
```{r}
ggplot(mko_clean, 
       aes(x = month_name, y = Temp_bot)) +
  geom_violin() +
  geom_boxplot(color = "gray", alpha = 0.5, width = 0.1,
               outlier.color = "black") +
  scale_x_discrete(limits = rev(month.name)) +
  coord_flip()


# half and half 
ggplot(data = penguins,
       aes(x = species, 
           y = bill_length_mm,
           fill = species)) +
  see::geom_violindot(size_dots = 5, alpha = 0.5)
```
# More charts!

## Line graphs 
- avoid spaghetti plots 
- cutting the y-axis
- adjusting aspect ratio 

## Area charts
- 1 vs. multiple groups
- how to interpret
- stacked vs proportional 

```{r}
library(tidycensus)
library(janitor)
library(gghighlight)

# load API key
source(here::here("week3", "KEYS.R"))
census_api_key(censusKEY)

lyme <- read_csv(here::here("week3", "data", "LD-Case-Counts-by-County-01-20.csv"))
```
## Wrangle
```{r}
# lyme data ----
lyme_clean <- lyme %>% 
  clean_names() %>% 
  rename(county = ctyname,
         state = stname,
         status = ststatus) %>% 
  pivot_longer(cols = 6:25,
               names_to = "city_year",
               values_to = "reported_cases") %>% 
  mutate(year = as.factor(str_remove(city_year, "cases")))

# calculate the total cases by state
lyme_by_state <- lyme_clean %>% 
  group_by(state) %>% 
  summarize(total_cases = sum(reported_cases))

us_state_pop <- get_estimates(geography = "state",
                              product = "population",
                              year = 2019) %>% 
  filter(variable == "POP") %>% 
  select(state = NAME, population = value)

lyme_pop <- left_join(lyme_by_state, us_state_pop) 
```

```{r}
#install.packages("geofacet") # facet data for different geographical regions using panels arranged in shapes that mimic geographic topology
library(geofacet)
library(tidyverse)

#..........................import data...........................
tuesdata <- tidytuesdayR::tt_load('2021-07-20')
drought <- tuesdata$drought

#..........................bad plots...........................
drought %>% 
  ggplot(aes(x = valid_start,
             y = drought_lvl)) +
  geom_boxplot() +
  geom_jitter()
```
```{r}
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                            wrangle drought data                          ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

drought_clean <- drought |>

  # select cols of interest & update names as needed ----
  select(date = valid_start, # make this more intuitive
         state_abb, 
         drought_lvl, 
         area_pct) |> 

  # add year, month & day cols using {lubridate} fxns ----
  mutate(year = year(date),
         month = month(date, label = TRUE, abbr = TRUE),
         day = day(date)) |>

  # add drought level conditions names ----
  mutate(drought_lvl_long = factor(drought_lvl,
                            levels = c("D4", "D3", "D2", "D1","D0", "None"),
                            labels = c("(D4) Exceptional", "(D3) Extreme",
                                       "(D2) Severe", "(D1) Moderate", "(D0) Abnormally Dry", 
                                       "No Drought"))) |>
  
  # reorder cols ----
  select(date, year, month, day, state_abb, drought_lvl, drought_lvl_long, area_pct)

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##       create stacked area plot of CA drought conditions through time     ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

drought_clean |> 
  
  # remove drought_lvl "None" & filter for just CA ----
  filter(drought_lvl != "None",
         state_abb == "CA") |> 
  
  # create ggplot ----
  ggplot(mapping = aes(x = date, y = area_pct, fill = drought_lvl_long)) +
  
  # reverse order of groups so level D4 is closest to x-axis ----
  geom_area(position = position_stack(reverse = TRUE)) +
  
  # update colors to match US Drought Monitor (colors selected using ColorPick Eyedropper from original USDM data viz) ----
  scale_fill_manual(values = c("#853904", "#FF0000", "#FFC100", "#FFD965", "#FFFF00")) +
  
  # set x-axis breaks & remove padding between data and x-axis ----
  scale_x_date(breaks = scales::breaks_pretty(n = 10),
               expand = c(0, 0)) +

  # set y-axis breaks & convert values to percentages & & remove padding between data and y-axis----
  scale_y_continuous(breaks = seq(0, 100, by = 10),
                     labels = scales::label_percent(scale = 1),
                     expand = c(0, 0)) +
  
  # add title ----
  labs(title = "Drought area in California") 

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##            BONUS: using {geofacet} to plot data for all states           ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#......remove HI & AK (no data) and also DC from preset grid.....
mygrid <- geofacet::us_state_grid1 |>
  filter(!code %in% c("DC", "HI", "AK"))

#..............................plot..............................

# NOTE: this takes a minute to render!

drought_clean |> 
  filter(drought_lvl != "None") |> 
  ggplot(aes(x = date, y = area_pct, fill = drought_lvl_long)) +
  geom_area(position = position_stack(reverse = TRUE)) +
  geofacet::facet_geo(~state_abb, grid = mygrid) +
  scale_fill_manual(values = c("#853904", "#FF0000", "#FFC100", "#FFD965", "#FFFF00"))
```

